Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-06-16 docs/paper.Rmd
write theoretical background (exposition and static q-q model)

>>>>

Becker’s early work introduced an economic approach to fertility, treating children analogous to “durable goods” that yield utility to parents but come at a cost \citep{doepke2015gary}. In Becker’s 1960 model, a household derives satisfaction from the number of children ($n$) and from the quality of each child ($q$), alongside conventional consumption of other goods ($y$). A simple representation is a utility function:

$$
U = U(n, q, y),
$$

with $U$ increasing in each argument up to some satiation point. Here quality $q$ can be thought of as the expenditure or investment per child (e.g. education spending, health care, nutrition), assumed for now to be the same for each child. Parents face a budget constraint that links quantity and quality: raising more children dilutes the resources available per child. A prototypical budget constraint (in static form) can be written as:

$$
p_y y + p_n n + p_q n q = I,
$$

where $I$ is total family income (or full income), $p_n$ represents baseline, non-discretionary costs associated with each additional child (e.g. expenditures on food, shelter, or clothing that are incurred irrespective of quality-enhancing investments), and $p_q$ denotes the marginal cost of investing in one unit of quality per child.The term $p_q n q$ captures total expenditure on quality for all children and is linear in $n$. As the number of children rises, parents must extend any chosen level of $q$ across a broader base, which amplifies the total cost of quality. Conversely, the term $p_n n$ implies that the cost of an additional child rises with the quality level $q$ already chosen, since each child must meet a higher standard of care or investment. For instance, a household that chooses to provide more education or better health care per child incurs an additional burden when it expands family size, as each child must receive the same enhanced level of investment. Similarly, a larger family increases the cumulative cost of quality, even if $q$ remains fixed, due to the need to replicate expenditures across more children. In short, the shadow price of child quality increases with $n$, and the shadow price of child quantity increases with $q$. The cost structure induces a mutual dependence between quantity and quality, such that any adjustment along one dimension alters the effective cost of the other.

Becker and Lewis (1973) formalize the mutually reinforcing nature of the quantity–quality cost structure. An increase in $n$ raises the total cost required to sustain a given level of $q$ for each child, while a higher level of $q$ raises the marginal cost associated with having an additional child. For example, allocating more resources to education or health per child increases the financial burden of expanding family size. This interdependence links the two decisions directly. The household cannot choose $n$ and $q$ in isolation; each choice alters the marginal cost of the other.

Mathematically, the trade-off appears in the first-order conditions of the household’s optimization problem. Let $\lambda$ represent the Lagrange multiplier on the full-income constraint.

$$
\mathcal{L} = U(n, q, y) + \lambda \left( I - p_y y - p_n n - p_q n q \right).
$$

The first-order conditions are:

$$
\frac{\partial \mathcal{L}}{\partial n} = U_n - \lambda(p_n + p_q q) = 0, \quad
\frac{\partial \mathcal{L}}{\partial q} = U_q - \lambda p_q n = 0, \quad
\frac{\partial \mathcal{L}}{\partial y} = U_y - \lambda p_y = 0.
$$

Combining the first two yields:

$$
\frac{U_n}{U_q} = \frac{p_n + p_q q}{p_q n}.
$$
This condition equates the marginal rate of substitution between quantity and quality to the ratio of their full marginal costs. The numerator rises with $q$, and the denominator rises with $n$. As one choice increases, the relative cost of the other becomes higher. This relationship induces substitution toward the less costly dimension. The trade-off between quantity and quality arises from the structure of the budget itself. It does not rely on specific assumptions about utility curvature or intrinsic substitutability \citep{becker1973interaction}.

This formulation implies two core predictions. Firstly, although both child quantity and child quality may rise with income, the household’s budget constraint can generate a negative relationship between income and fertility. As income increases, total spending on children tends to rise, but the allocation often favors quality over quantity. Becker illustrated this with the analogy of durable goods: wealthier households tend to upgrade the quality of a house or a car rather than acquire additional units. In a similar way, higher-income families often direct additional resources toward education, nutrition, or health per child. Within the model, an income increase ($\mathrm{d}I > 0$) produces a direct effect that makes children more affordable and an indirect effect that discourages fertility. As $q$ rises, the shadow price of an additional child also rises. If the marginal utility from higher quality exceeds that from larger family size, then the substitution effect outweighs the income effect, leading to a lower optimal $n$. This mechanism offers a structural explanation for the demographic transition: fertility tends to fall as households become richer, even when preferences remain unchanged.

Furthermore, a similar logic applies to changes in the cost parameters $p_q$ and $p_n$. A decline in $p_q$, such as through a policy that lowers the price of education or health care, increases $q$ and raises the marginal cost of quantity. This effect reduces optimal fertility. A rise in $p_n$, which may reflect higher child-rearing costs or a greater opportunity cost of parental time, reduces the appeal of larger families and can shift resources toward child quality. These outcomes follow from the structure of the budget constraint, without requiring any explicit preference for quality over quantity. Becker and Lewis noted that these comparative static results align with observed patterns. For example, increases in women’s wages often reduce fertility more than they reduce educational spending per child. This asymmetry reflects the model’s central feature: quantity and quality are linked through their cost structure. An increase in one raises the marginal cost of the other. The model explains how households make trade-offs between the number of children and investments in each.

While the early Q–Q models were static (one-period) representations, subsequent contributions extended the framework to consider fertility and child investment over multiple periods or even multiple generations. A central development in this literature was the incorporation of intergenerational human capital dynamics, where parents derive utility not only from the number and quality of children in the present, but also from the long-run outcomes of their offspring. These extensions allowed child quality to evolve endogenously across time, rather than being determined solely within a single period.

### 2025-06-16 docs/paper.tex
write theoretical background (exposition and static q-q model)

The trade-off between child quantity and child quality is a foundational
concept in the economics of the family. First articulated by
\citet{becker1960economic} and extended in subsequent models of
household behavior \citep{beckerlewis1973, beckertomes1976}, this
framework posits that parents allocate finite resources---both financial
and non-financial---across children. An increase in fertility reduces
the resources available per child and, under binding constraints, may
lead to lower investments in health, education, and other forms of human
capital. This mechanism has served as an explanatory model for changes
in fertility behavior and the evolution of population structures in low-
and middle-income countries.

@@

The analysis proceeds in three stages. I first replicate existing work
\citep{dumas2010fertility} to confirm that the contraceptive ban led to
an increase in fertility among affected women. I then estimate
reduced-form effects of policy exposure on nutritional outcomes for
children under five years of age. Finally, I examine heterogeneity in
effects across subsamples defined by maternal education, household
wealth, and access to prenatal care. These dimensions serve as proxies
for household resource availability and capacity to buffer the
nutritional consequences of fertility increases.

This research contributes to the literature in several respects. It
introduces new evidence on the nutritional consequences of fertility
shocks in an urban developing-country context. It expands the
measurement of child quality beyond educational outcomes to include
biological indicators. It also adds to the limited body of work that
uses plausibly exogenous variation in reproductive health policy to
study downstream effects on human capital. More broadly, the findings
underscore the potential for local family planning restrictions to
generate unintended consequences for child well-being in settings
characterized by poverty, food insecurity, and institutional fragility.

>>>>

The trade-off between child quantity and child quality is a foundational
concept in the economics of the family. First articulated by
\citet{becker1960economic} and extended in subsequent models of
household behavior \citep{becker1973interaction, becker1976child}, this
framework posits that parents allocate finite resources---both financial
and non-financial---across children. An increase in fertility reduces
the resources available per child and, under binding constraints, may
lead to lower investments in health, education, and other forms of human
capital. This mechanism has served as an explanatory model for changes
in fertility behavior and the evolution of population structures in low-
and middle-income countries.

@@

The analysis proceeds in three stages. I first replicate existing work
\citep{dumas2019sex} to confirm that the contraceptive ban led to an
increase in fertility among affected women. I then estimate reduced-form
effects of policy exposure on nutritional outcomes for children under
five years of age. Finally, I examine heterogeneity in effects across
subsamples defined by maternal education, household wealth, and access
to prenatal care. These dimensions serve as proxies for household
resource availability and capacity to buffer the nutritional
consequences of fertility increases.

This study contributes to the literature in several important ways. It
provides new evidence on how increases in fertility---caused by policy
restrictions on family planning---can affect child nutrition in poor,
urban communities. Most past research has focused on education, but this
study expands the idea of child quality to include biological measures
such as stunting and wasting. It also adds to the small number of
studies that use unexpected changes in reproductive health policy to
examine long-term effects on children's well-being. More broadly, the
results show that local restrictions on family planning can
unintentionally harm children's health, especially in settings where
families already face poverty, food insecurity, and limited public
services.

\section{Review of Related Literature}

The quantity--quality (Q--Q) theory, a central idea in modern family
economics, holds that parents face a trade-off between the number of
children and the ``quality'' of investment---such as education or
health---they can provide to each. Quality in this context refers to the
human capital of each child: attributes like education, health, and
nutrition that enhance a child's future productivity and well-being. The
genesis of this idea traces back to Gary Becker's seminal work around
1960, which for the first time treated children as economic goods
subject to parental choice and budget constraints
\citep{becker1960economic}. Becker argued that as families become
wealthier, they may not simply want more children, but rather
better-raised children, much as a household might prefer a
higher-quality car or house over a greater quantity of them. This
proposition led to a formal theory in which increases in income or
changes in economic conditions cause parents to substitute child quality
for quantity, consistent with historical patterns of lower fertility and
higher educational attainment during economic development
\citep{galor2000population}.

In what follows, I review the theoretical foundations of the Q-Q model
and its evolution in the literature. I begin with the static models of
Becker \citep{becker1960economic} and Becker--Lewis
\citep{becker1973interaction}, which first formalized the trade-off
within a household utility maximization framework. We then examine
extensions to dynamic, intergenerational settings, including the
contributions of Becker and Tomes \citep{becker1976child} on child
endowments and the altruistic dynastic model associated with Barro and
Becker \citep{barro1989fertility}. Next, I turn to macroeconomic and
unified growth models, notably Galor and Weil
\citep{galor2000population} and Galor and Moav \citep{galor2002natural},
which integrate the Q--Q mechanism into a general theory of demographic
and economic transformation.

Finally, I discuss more recent refinements that enrich the basic model
by incorporating credit constraints \citep{doepke2004accounting},
intra-household bargaining \citep{doepke2019bargaining}, and
multi-dimensional child quality
\citep{hoddinott2013economic, kalemli2002does}, with a special emphasis
on health and nutrition. My focus is on how the Q--Q framework has been
applied to understand fertility and child investment patterns,
especially in developing country contexts where resource constraints and
health outcomes are paramount.

\subsection{Theoretical Background}

Becker's early work introduced an economic approach to fertility,
treating children analogous to ``durable goods'' that yield utility to
parents but come at a cost \citep{doepke2015gary}. In Becker's 1960
model, a household derives satisfaction from the number of children
(\(n\)) and from the quality of each child (\(q\)), alongside
conventional consumption of other goods (\(y\)). A simple representation
is a utility function:

\[
U = U(n, q, y),
\]

with \(U\) increasing in each argument up to some satiation point. Here
quality \(q\) can be thought of as the expenditure or investment per
child (e.g.~education spending, health care, nutrition), assumed for now
to be the same for each child. Parents face a budget constraint that
links quantity and quality: raising more children dilutes the resources
available per child. A prototypical budget constraint (in static form)
can be written as:

\[
p_y y + p_n n + p_q n q = I,
\]

where \(I\) is total family income (or full income), \(p_n\) represents
baseline, non-discretionary costs associated with each additional child
(e.g.~expenditures on food, shelter, or clothing that are incurred
irrespective of quality-enhancing investments), and \(p_q\) denotes the
marginal cost of investing in one unit of quality per child.The term
\(p_q n q\) captures total expenditure on quality for all children and
is linear in \(n\). As the number of children rises, parents must extend
any chosen level of \(q\) across a broader base, which amplifies the
total cost of quality. Conversely, the term \(p_n n\) implies that the
cost of an additional child rises with the quality level \(q\) already
chosen, since each child must meet a higher standard of care or
investment. For instance, a household that chooses to provide more
education or better health care per child incurs an additional burden
when it expands family size, as each child must receive the same
enhanced level of investment. Similarly, a larger family increases the
cumulative cost of quality, even if \(q\) remains fixed, due to the need
to replicate expenditures across more children. In short, the shadow
price of child quality increases with \(n\), and the shadow price of
child quantity increases with \(q\). The cost structure induces a mutual
dependence between quantity and quality, such that any adjustment along
one dimension alters the effective cost of the other.

Becker and Lewis (1973) formalize the mutually reinforcing nature of the
quantity--quality cost structure. An increase in \(n\) raises the total
cost required to sustain a given level of \(q\) for each child, while a
higher level of \(q\) raises the marginal cost associated with having an
additional child. For example, allocating more resources to education or
health per child increases the financial burden of expanding family
size. This interdependence links the two decisions directly. The
household cannot choose \(n\) and \(q\) in isolation; each choice alters
the marginal cost of the other.

Mathematically, the trade-off appears in the first-order conditions of
the household's optimization problem. Let \(\lambda\) represent the
Lagrange multiplier on the full-income constraint.

\[
\mathcal{L} = U(n, q, y) + \lambda \left( I - p_y y - p_n n - p_q n q \right).
\]

The first-order conditions are:

\[
\frac{\partial \mathcal{L}}{\partial n} = U_n - \lambda(p_n + p_q q) = 0, \quad
\frac{\partial \mathcal{L}}{\partial q} = U_q - \lambda p_q n = 0, \quad
\frac{\partial \mathcal{L}}{\partial y} = U_y - \lambda p_y = 0.
\]

Combining the first two yields:

\[
\frac{U_n}{U_q} = \frac{p_n + p_q q}{p_q n}.
\] This condition equates the marginal rate of substitution between
quantity and quality to the ratio of their full marginal costs. The
numerator rises with \(q\), and the denominator rises with \(n\). As one
choice increases, the relative cost of the other becomes higher. This
relationship induces substitution toward the less costly dimension. The
trade-off between quantity and quality arises from the structure of the
budget itself. It does not rely on specific assumptions about utility
curvature or intrinsic substitutability \citep{becker1973interaction}.

This formulation implies two core predictions. Firstly, although both
child quantity and child quality may rise with income, the household's
budget constraint can generate a negative relationship between income
and fertility. As income increases, total spending on children tends to
rise, but the allocation often favors quality over quantity. Becker
illustrated this with the analogy of durable goods: wealthier households
tend to upgrade the quality of a house or a car rather than acquire
additional units. In a similar way, higher-income families often direct
additional resources toward education, nutrition, or health per child.
Within the model, an income increase (\(\mathrm{d}I > 0\)) produces a
direct effect that makes children more affordable and an indirect effect
that discourages fertility. As \(q\) rises, the shadow price of an
additional child also rises. If the marginal utility from higher quality
exceeds that from larger family size, then the substitution effect
outweighs the income effect, leading to a lower optimal \(n\). This
mechanism offers a structural explanation for the demographic
transition: fertility tends to fall as households become richer, even
when preferences remain unchanged.

Furthermore, a similar logic applies to changes in the cost parameters
\(p_q\) and \(p_n\). A decline in \(p_q\), such as through a policy that
lowers the price of education or health care, increases \(q\) and raises
the marginal cost of quantity. This effect reduces optimal fertility. A
rise in \(p_n\), which may reflect higher child-rearing costs or a
greater opportunity cost of parental time, reduces the appeal of larger
families and can shift resources toward child quality. These outcomes
follow from the structure of the budget constraint, without requiring
any explicit preference for quality over quantity. Becker and Lewis
noted that these comparative static results align with observed
patterns. For example, increases in women's wages often reduce fertility
more than they reduce educational spending per child. This asymmetry
reflects the model's central feature: quantity and quality are linked
through their cost structure. An increase in one raises the marginal
cost of the other. The model explains how households make trade-offs
between the number of children and investments in each.

While the early Q--Q models were static (one-period) representations,
subsequent contributions extended the framework to consider fertility
and child investment over multiple periods or even multiple generations.
A central development in this literature was the incorporation of
intergenerational human capital dynamics, where parents derive utility
not only from the number and quality of children in the present, but
also from the long-run outcomes of their offspring. These extensions
allowed child quality to evolve endogenously across time, rather than
being determined solely within a single period.
